---
title: "HW1"
author: "Fan Hong, Dantong Huang, Xin Liu, and Hao Wang (in alphabetical order)"
date: "January 23, 2017"
output: html_document
---

Most of codes are from Rob's notes "knn-bvt.pdf" and the cross validation function is from "https://raw.githubusercontent.com/ChicagoBoothML/HelpR/master".

## Problem 5.1
Summary of problem 5.1:

1) By "eye-ball" method we guessed a good k=50
2) By 5-fold cross-validatifon (10 times) we obtained the best k=22
3) We found kNN with CV has slightly better performance (in terms of covariance with true labels) than kNN with "eyeball": 0.8768632 v.s 0.8695700
4) We made a prediction for a used car with 100,000 miles with k=22 from CV and its price is 18114.82

Please check the attached code for these results:  
```{r Clean0, include=FALSE}
rm(list=ls())
```
Load library:
```{r LL0}
library(kknn)
library(data.table)
source('docv.R')
```

Load data:
```{r LD0}
UsedCar_small <- fread("UsedCars_small.csv",header = T,stringsAsFactors = F)
UsedCarData = UsedCar_small
```

Do feature selection (only "mileage") and data scaling:
```{r FSDC0}
x = cbind(UsedCarData$mileage)
colnames(x) = c("mileage")
y = UsedCarData$price
mmsc=function(x) {return((x-min(x))/(max(x)-min(x)))}
xs = apply(x,2,mmsc) #apply scaling function to each column of x
```

“Eye-ball” guess:
```{r, echo=FALSE}
par(mfrow=c(2,2))
k=10
kf = kknn(y~x,data.frame(y,x),data.frame(x=sort(x)),k=k,kernel = "rectangular")
plot(x,y,cex.lab=1.2,main=paste("k=",toString(k)),xlab="mileage",ylab="price")
lines(sort(x),kf$fitted,col="red",lwd=2,cex.lab=2)
k=30
kf = kknn(y~x,data.frame(y,x),data.frame(x=sort(x)),k=k,kernel = "rectangular")
plot(x,y,cex.lab=1.2,main=paste("k=",toString(k)),xlab="mileage",ylab="price")
lines(sort(x),kf$fitted,col="red",lwd=2,cex.lab=2)
k=50
kf = kknn(y~x,data.frame(y,x),data.frame(x=sort(x)),k=k,kernel = "rectangular")
plot(x,y,cex.lab=1.2,main=paste("k=",toString(k)),xlab="mileage",ylab="price")
lines(sort(x),kf$fitted,col="red",lwd=2,cex.lab=2)
k=70
kf = kknn(y~x,data.frame(y,x),data.frame(x=sort(x)),k=k,kernel = "rectangular")
plot(x,y,cex.lab=1.2,main=paste("k=",toString(k)),xlab="mileage",ylab="price")
lines(sort(x),kf$fitted,col="red",lwd=2,cex.lab=2)
```

Do 5-fold cross validation (10 times) and find the best k:
```{r CV0}
kv = 2:100
set.seed(99)
cvmean = rep(0,length(kv)) #will keep average rmse here
ndocv = 10 #number of CV splits to try
n=length(y)
cvmat = matrix(0,length(kv),ndocv) #keep results for each split
for(i in 1:ndocv) {
  cvtemp = docvknn(xs,y,kv,nfold=5,verbose=F)
  cvmean = cvmean + cvtemp
  cvmat[,i] = sqrt(cvtemp/n)
}
cvmean = cvmean/ndocv
cvmean = sqrt(cvmean/n)
Bestk = which.min(cvmean) + kv[1] - 1
cat("Best k is: ", Bestk)
```


Fit using all the data with the best k from CV and compared with kNN with eyeball guess:
```{r FitAllData0}

par(mfrow=c(1,2))
eyeballk = 50
kf = kknn(y~x,data.frame(y,x),data.frame(x=sort(x)),k=eyeballk,kernel = "rectangular")
plot(x,y,cex.lab=1.2,main=paste("Eyeball k=",toString(eyeballk)),xlab="mileage",ylab="price")
lines(sort(x),kf$fitted,col="red",lwd=2,cex.lab=2)
k=Bestk
kf = kknn(y~x,data.frame(y,x),data.frame(x=sort(x)),k=Bestk,kernel = "rectangular")
plot(x,y,cex.lab=1.2,main=paste("CV k=",toString(k)),xlab="mileage",ylab="price")
lines(sort(x),kf$fitted,col="red",lwd=2,cex.lab=2)

ddf = data.frame(y,xs)
nearBestk5_1 = kknn(y~.,ddf,ddf,k=Bestk,kernel = "rectangular")
neareyeball = kknn(y~.,ddf,ddf,k=eyeballk,kernel = "rectangular")
fmat = cbind(y,nearBestk5_1$fitted,neareyeball$fitted)
colnames(fmat)=c("y",paste("kNN CV",toString(Bestk)),paste("kNN Eyeball",toString(eyeballk)))
pairs(fmat)
print(cor(fmat))
```
We see that kNN with k from cross validation has slightly better performance than kNN with k from "eyeball".

Predict the price of a used car with 100,000 miles:
```{r Pred100000}
x1=100000
x1s = (x1-min(x[,1]))/(max(x[,1])-min(x[,1]))
near = kknn(y~.,data.frame(y,xs),data.frame(mileage=x1s),k=Bestk,kernel = "rectangular")
cat("knn predicted value: ",near$fitted,"\n")
```

\newpage

## Problem 6.1
Summary of problem 6.1:

1) By 10-fold cross-validatifon (10 times) we obtained the best k=28
2) We made a prediction for a used car with 100,000 miles and 2008 year with k=28 and its price is 31596.43
3) We found kNN with ("mileages"", "year") has much better performance (in terms of covariance with true labels) than kNN with only "mileages": 0.9578319 v.s. 0.8768632  

Please check the attached code for these results:  

Do feature selection ("mileage" and "year") and data scaling: 
```{r FSDC}
x = cbind(UsedCarData$mileage,UsedCarData$year)
colnames(x) = c("mileage","year")
y = UsedCarData$price
mmsc=function(x) {return((x-min(x))/(max(x)-min(x)))}
xs = apply(x,2,mmsc) 
```
<!--
Plot price v.s. mileage and price v.s. year:
```{r price v.s. feature}
par(mfrow=c(1,2))
plot(x[,1],y,xlab="mileage",ylab="price")
plot(x[,2],y,xlab="year",ylab="price")
```
-->

Run cross validation and find the best k:
```{r CV}
kv = 2:100 
set.seed(99)
cvmean = rep(0,length(kv)) #will keep average rmse here
ndocv = 10 #number of CV splits to try
n=length(y)
cvmat = matrix(0,length(kv),ndocv) #keep results for each split
for(i in 1:ndocv) {
  cvtemp = docvknn(xs,y,kv,nfold=10,verbose=F)
  cvmean = cvmean + cvtemp
  cvmat[,i] = sqrt(cvtemp/n)
}
cvmean = cvmean/ndocv
cvmean = sqrt(cvmean/n)
plot(kv,cvmean,type="n",ylim=range(cvmat),xlab="k",cex.lab=1.5)
for(i in 1:ndocv) lines(kv,cvmat[,i],col=i,lty=3) 
lines(kv,cvmean,type="b",col="black",lwd=3) 
Bestk = which.min(cvmean) + kv[1] - 1
cat("Best k is: ", Bestk, "\n")
```

Predict the price of a used car with 75,000 miles, year 2008
```{r Predit}
ddf = data.frame(y,xs)
x1=75000; x2=2008
x1s = (x1-min(x[,1]))/(max(x[,1])-min(x[,1]))
x2s = (x2-min(x[,2]))/(max(x[,2])-min(x[,2]))
near = kknn(y~.,ddf,data.frame(mileage=x1s,year=x2s),k=Bestk,kernel = "rectangular")
cat("knn predicted value: ",near$fitted,"\n")
```

Compare kNN (k=28) with (mile,year) with kNN (k=22) with mile:
```{r FitAllData}
nearBestk6_1 = kknn(y~.,ddf,ddf,k=Bestk,kernel = "rectangular")
fmat = cbind(y,nearBestk6_1$fitted,nearBestk5_1$fitted)
colnames(fmat)=c("y", "kNN with miles and year","kNN with miles")
pairs(fmat)
print(cor(fmat))
```